Modeling Outbreak Frequency by Season and Pathogen: Insights and Predictions for 2025

### 2024-12-02 paper/paper.qmd
Add files via upload

---
title: "Modeling Outbreak Frequency by Season and Pathogen: Insights and Predictions for 2025"
subtitle: "Seasonal Patterns and Stable COVID-19 Outbreaks Across Healthcare Settings"
author: 
  - Tianrui Fu
thanks: "Code and data are available at: [https://github.com/FrankFU323/Toronto-Healthcare-outbreak-trends.git](https://github.com/FrankFU323/Toronto-Healthcare-outbreak-trends.git)."
date: today
date-format: long
abstract: "We analyze the outbreak dynamics of respiratory pathogens in Toronto’s healthcare settings, focusing on seasonal trends, pathogen-specific patterns, and predictive modeling. Using Poisson regression on historical data, we find COVID-19 dominates outbreak frequencies, with summer months showing peaks across most pathogens and lower activity in winter. Predictions for 2025 indicate similar seasonal patterns, reinforcing the importance of seasonally adjusted preparedness measures. These results highlight the need for strategic resource allocation to mitigate the burden of outbreaks while ensuring healthcare systems remain resilient to seasonal variability."
format: pdf
number-sections: true
bibliography: references.bib
toc: true
---

@@

```{r, fig.pos= 'H', fig.width=10, fig.height=6}
#| echo: false
library(arrow)
data <- read_parquet(here::here("data/02-analysis_data/analysis_data.parquet"))

```{r, fig.pos= 'H', fig.width=10, fig.height=6}
#| label: fig-count
#| fig-cap: Outbreak Counts by Pathogen and Season
#| echo: false

@@

# Create the bar plot
ggplot(plot_data, aes(x = causative_agent_1, y = total_outbreaks, fill = season)) +
  geom_bar(stat = "identity", position = "dodge") + # Dodge for season-wise bars
  geom_text(aes(label = total_outbreaks), 
            position = position_dodge(width = 0.9), 
            vjust = -0.5, size = 3) + # Annotate with outbreak counts
  labs(
    x = "Pathogen",
    y = "Total Outbreak Count",
    fill = "Season"
  ) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1) # Rotate x-axis labels
  )

@@

```{r, fig.pos= 'H', fig.width=10, fig.height=6}
#| label: fig-agent
#| fig-cap: Outbreak Counts by Pathogen
#| echo: false
#| warning: false

ggplot(data, aes(x = causative_agent_1, fill = causative_agent_1)) +
  geom_bar() +
  geom_text(stat = 'count', aes(label = ..count..), vjust = -0.5) +
  labs(
    x = "Pathogen",
    y = "Total Count"
  ) +
  theme_minimal() +
  theme(legend.position = "none")

@@

```{r, fig.pos= 'H', fig.width=10, fig.height=6}
#| label: fig-season
#| fig-cap: Outbreak Counts by Season
#| echo: false

ggplot(data, aes(x = season, fill = season)) +
  geom_bar() +
  geom_text(stat = 'count', aes(label = ..count..), vjust = -0.5) +
  labs(
    x = "Season",
    y = "Total Count"
  ) +
  theme_minimal() +
  theme(legend.position = "none")

```
### Month

@@

```{r, fig.pos= 'H', fig.width=10, fig.height=6}
#| label: fig-month
#| fig-cap: Monthly Outbreak Trends
#| echo: false

ggplot(data, aes(x = factor(month, levels = 1:12), fill = factor(month))) +
  geom_bar() +
  geom_text(stat = 'count', aes(label = ..count..), vjust = -0.5) +
  scale_x_discrete(labels = month.abb) +
  labs(
    x = "Month",
    y = "Total Count"
  ) +
  theme_minimal() +
  theme(legend.position = "none")

@@

```{r, fig.pos= 'H', fig.width=10, fig.height=6}
#| label: fig-outbreak_setting
#| fig-cap: Outbreak Count by Setting
#| echo: false

ggplot(data, aes(x = outbreak_setting, fill = outbreak_setting)) +
  geom_bar() +
  geom_text(stat = 'count', aes(label = ..count..), vjust = -0.5) +
  labs(
    x = "Setting",
    y = "Total Count"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  theme(legend.position = "none")

@@

## Model justification

The Poisson regression model was chosen for its suitability in handling count data, such as outbreak frequencies in healthcare settings, where the dependent variable `outbreak_count` reflects non-negative integers. This aligns naturally with the properties of the Poisson distribution. The model captures the relationship between the frequency of outbreaks and key predictors, including `causative_agent_1`, `month`, `season`, and `outbreak_setting`, which were identified as relevant through a thorough examination of the dataset. These variables represent temporal, environmental, and pathogen-specific factors influencing outbreak patterns.

@@

@@

@@

Additionally, in @sec-model_prediction_results, the model was used to predict potential outbreak trends for 2025. These predictions assume consistent patterns in the predictor variables and similar environmental conditions to those observed in 2024. However, the predictions are limited by the exclusion of other potential influencing factors, such as changes in public health interventions, population dynamics, or pathogen evolution. The more particular information see in @sec-discussion. The forecast aims to provide an estimate of outbreak patterns under comparable conditions, serving as a baseline scenario to guide preparedness and response planning.

@@

modelsummary(
  model = poisson_season_time,
  output = "kableExtra", 
  stars = FALSE
) %>%
  kable_styling(
    full_width = FALSE, 
    position = "center", 
    latex_options = c("hold_position", "scale_down")
  ) %>%
  column_spec(column = 1, width = "14em") %>% 
  column_spec(column = 2, width = "14em") %>%
  add_header_above(c(" " = 1, "Model Summary" = 1))

```
\newpage

@@

b <- list(geom_vline(xintercept = 0, color = 'orange'),
          annotate("rect", alpha = .1,
                   xmin = -.5, xmax = .5, 
                   ymin = -Inf, ymax = Inf),
          geom_point(aes(y = term, x = estimate), alpha = .3, 
                     size = 8, color = 'red'))

@@

ggplot(data, aes(x = month, y = predicted, color = causative_agent_1)) +
  geom_rect(data = season_data, aes(xmin = start_month, xmax = end_month, ymin = ymin, ymax = ymax, fill = season),
            inherit.aes = FALSE, alpha = 0.2) + 
  geom_line() +
  geom_smooth(method = "loess", se = FALSE) +
  facet_wrap(~ causative_agent_1, scales = "free_y") +
  labs(
    x = "Month",
    y = "Predicted Outbreak Count",
    fill = "Season"
  ) +
  scale_x_continuous(breaks = 1:12, labels = month.abb) +
  scale_fill_manual(values = c("Winter" = "blue", "Spring" = "green", "Summer" = "yellow", "Fall" = "orange")) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 60, hjust = 1)
  )
```

@@

# Create the plot
ggplot(new_data, aes(x = month, y = predicted_outbreak, color = causative_agent_1)) +
  geom_rect(data = season_data, aes(xmin = start_month, xmax = end_month, ymin = ymin, ymax = ymax, fill = season),
            inherit.aes = FALSE, alpha = 0.2) +
  geom_line() +
  geom_smooth(method = "loess", se = FALSE) +
  facet_wrap(~ causative_agent_1, scales = "free_y") +
  labs(
    x = "Month",
    y = "Predicted Outbreak Count",
    fill = "Season",
    color = "Pathogen"
  ) +
  scale_x_continuous(breaks = 1:12, labels = month.abb) +
  scale_fill_manual(values = c("Winter" = "blue", "Spring" = "green", "Summer" = "yellow", "Fall" = "orange")) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 60, hjust = 1)
  )

@@

ggplot(combined_data, aes(x = month, y = count, color = source)) +
  geom_line(size = 1) +
  facet_wrap(~ causative_agent_1, scales = "free_y") +
  labs(
    x = "Month",
    y = "Outbreak Count",
    color = "Source"
  ) +
  scale_x_continuous(breaks = 1:12, labels = month.abb) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 60, hjust = 1),
    legend.position = "bottom"
  )
```

@@

@@

par(mfrow = c(2,3))
plot(poisson_season_time, which = 1:6)
```

>>>>

---
title: "Modeling Outbreak Frequency by Season and Pathogen: Insights and Predictions for 2025"
subtitle: "COVID-19 Dominates with Steady Frequency, Other Pathogens Peak in Summer"
author: 
  - Tianrui Fu
thanks: "Code and data are available at: [https://github.com/FrankFU323/Toronto-Healthcare-outbreak-trends.git](https://github.com/FrankFU323/Toronto-Healthcare-outbreak-trends.git)."
date: today
date-format: long
abstract: "We analyze the outbreak dynamics of respiratory pathogens in Toronto’s healthcare settings, focusing on seasonal trends, pathogen-specific patterns, and predictive modeling. Using Poisson regression on historical data, we find COVID-19 dominates outbreak frequencies, with summer months showing peaks across most pathogens and lower activity in winter. Predictions for 2025 indicate similar seasonal patterns, reinforcing the importance of seasonally adjusted preparedness measures. These results highlight the need for strategic resource allocation to mitigate the burden of outbreaks while ensuring healthcare systems remain resilient to seasonal variability."
format: pdf
number-sections: true
bibliography: references.bib
toc: true
---

@@

The @fig-count provides a more detailed and insightful view of the outcome variable `outbreak_count` by breaking it down into specific groups such as `causative_agent_1` and `season`. This grouping allows us to explore how outbreaks vary across different pathogens and seasons, providing a richer understanding of patterns within the data. The plot with just `outbreak_count` is included in [Appendix -@sec-plot-details].

```{r, fig.pos= 'H', fig.width=10, fig.height=5}
#| echo: false
library(arrow)
data <- read_parquet(here::here("data/02-analysis_data/analysis_data.parquet"))
```

```{r, fig.pos= 'H', fig.width=10, fig.height=5}
#| label: fig-count
#| fig-cap: Outbreak Counts by Pathogen and Season
#| echo: false

@@

# Create the bar plot
ggplot(plot_data, aes(x = causative_agent_1, y = total_outbreaks, fill = season)) +
  geom_bar(stat = "identity", position = "dodge") + # Dodge for season-wise bars
  geom_text(aes(label = total_outbreaks),
    position = position_dodge(width = 0.9),
    vjust = -0.5, size = 3
  ) + # Annotate with outbreak counts
  labs(
    x = "Pathogen",
    y = "Total Outbreak Count",
    fill = "Season"
  ) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 60, hjust = 1) # Rotate x-axis labels
  )
```

@@

```{r, fig.pos= 'H', fig.width=10, fig.height=5}
#| label: fig-agent
#| fig-cap: Outbreak Counts by Pathogen
#| echo: false
#| warning: false

ggplot(data, aes(x = causative_agent_1, fill = causative_agent_1)) +
  geom_bar() +
  geom_text(stat = "count", aes(label = ..count..), vjust = -0.5) +
  labs(
    x = "Pathogen",
    y = "Total Count"
  ) +
  theme_minimal() +
  theme(legend.position = "none")
```

@@

```{r, fig.pos= 'H', fig.width=10, fig.height=5}
#| label: fig-season
#| fig-cap: Outbreak Counts by Season
#| echo: false

ggplot(data, aes(x = season, fill = season)) +
  geom_bar() +
  geom_text(stat = "count", aes(label = ..count..), vjust = -0.5) +
  labs(
    x = "Season",
    y = "Total Count"
  ) +
  theme_minimal() +
  theme(legend.position = "none")
```
### Month

@@

```{r, fig.pos= 'H', fig.width=10, fig.height=5}
#| label: fig-month
#| fig-cap: Monthly Outbreak Trends
#| echo: false

ggplot(data, aes(x = factor(month, levels = 1:12), fill = factor(month))) +
  geom_bar() +
  geom_text(stat = "count", aes(label = ..count..), vjust = -0.5) +
  scale_x_discrete(labels = month.abb) +
  labs(
    x = "Month",
    y = "Total Count"
  ) +
  theme_minimal() +
  theme(legend.position = "none")
```

@@

```{r, fig.pos= 'H', fig.width=10, fig.height=5}
#| label: fig-outbreak_setting
#| fig-cap: Outbreak Count by Setting
#| echo: false

ggplot(data, aes(x = outbreak_setting, fill = outbreak_setting)) +
  geom_bar() +
  geom_text(stat = "count", aes(label = ..count..), vjust = -0.5) +
  labs(
    x = "Setting",
    y = "Total Count"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  theme(legend.position = "none")
```

@@

## Model justification
The Poisson regression model was chosen for its suitability in handling count data, such as outbreak frequencies in healthcare settings, where the dependent variable `outbreak_count` reflects non-negative integers. This aligns naturally with the properties of the Poisson distribution. The model captures the relationship between the frequency of outbreaks and key predictors, including `causative_agent_1`, `month`, `season`, and `outbreak_setting`, which were identified as relevant through a thorough examination of the dataset. These variables represent temporal, environmental, and pathogen-specific factors influencing outbreak patterns.

@@

To further evaluate the model's adequacy, a Reduced Model was created through variable selection by systematically excluding non-significant predictors to simplify the framework. However, comparison results indicated in @tbl-modelcomparison that the Poisson Season Time Model outperformed the Reduced Model, evidenced by a lower AIC (449.29 vs. 508.79) and RMSE (2.60 vs. 4.43). These findings underscore the importance of retaining interaction terms like `month * season` and pathogen-specific variables to effectively capture temporal and outbreak-specific dynamics. The Poisson Season Time Model was thus deemed the most robust and informative framework for understanding outbreak patterns.

@@

```{r,fig.pos= 'H'}
#| echo: false
#| eval: true
#| label: tbl-modelcomparison
#| tbl-cap: Comparison of full and reduced model
#| warning: false

poisson_season_time <- readRDS(file = here::here("models/poisson_season_time.rds"))
reduced_model <- readRDS(file = here::here("models/reduced_selected_model.rds"))
# Compare Full Model and reduced Model

# Summaries of the models
invisible(summary(poisson_season_time))
invisible(summary(reduced_model))

# AIC and BIC comparison
aic_full <- round(AIC(poisson_season_time), 2)
bic_full <- round(BIC(poisson_season_time), 2)
aic_reduced <- round(AIC(reduced_model), 2)
bic_reduced <- round(BIC(reduced_model), 2)

# RMSE Calculation
rmse <- function(actual, predicted) {
  round(sqrt(mean((actual - predicted)^2)), 2)
}

# Predicted values
full_predictions <- predict(poisson_season_time, type = "response")
reduced_predictions <- predict(reduced_model, type = "response")

# Calculate RMSE for both models
rmse_full <- rmse(data$outbreak_count, full_predictions)
rmse_reduced <- rmse(data$outbreak_count, reduced_predictions)

# Likelihood Ratio Test
lrt <- anova(reduced_model, poisson_season_time, test = "LRT")

comparison <- data.frame(
  Metric = c("AIC", "BIC", "RMSE", "Likelihood Ratio Test (p-value)"),
  `Full Model` = c(aic_full, bic_full, rmse_full, lrt$`Pr(>Chi)`[2]),
  `Reduced Model` = c(aic_reduced, bic_reduced, rmse_reduced, "N/A") # LRT only tests nested models
)

# Print the table
comparison %>%
  kbl() %>%
  kable_styling(full_width = FALSE, bootstrap_options = c("striped", "hover", "condensed"))
```

@@

Additionally, in @sec-model_prediction_results, the model was used to predict potential outbreak trends for 2025. These predictions assume consistent patterns in the predictor variables and similar environmental conditions to those observed in 2024. However, the predictions are limited by the exclusion of other potential influencing factors, such as changes in public health interventions, population dynamics, or pathogen evolution. The more particular information see in @sec-discussion.

@@

modelsummary(
  model = poisson_season_time,
  output = "kableExtra",
  stars = FALSE
) %>%
  kable_styling(
    full_width = FALSE,
    position = "center",
    latex_options = c("hold_position", "scale_down"),
  ) %>%
  column_spec(column = 1, width = "14em") %>%
  column_spec(column = 2, width = "14em") %>%
  add_header_above(c(" " = 1, "Model Summary" = 1))
```
\newpage

@@

b <- list(
  geom_vline(xintercept = 0, color = "orange"),
  annotate("rect",
    alpha = .1,
    xmin = -.5, xmax = .5,
    ymin = -Inf, ymax = Inf
  ),
  geom_point(aes(y = term, x = estimate),
    alpha = .3,
    size = 8, color = "red"
  )
)

@@

ggplot(data, aes(x = month, y = predicted, color = causative_agent_1)) +
  geom_rect(
    data = season_data, aes(xmin = start_month, xmax = end_month, ymin = ymin, ymax = ymax, fill = season),
    inherit.aes = FALSE, alpha = 0.2
  ) +
  geom_line() +
  geom_smooth(method = "loess", se = FALSE) +
  facet_wrap(~causative_agent_1, scales = "free_y") +
  labs(
    x = "Month",
    y = "Predicted Outbreak Count",
    fill = "Season"
  ) +
  scale_x_continuous(breaks = 1:12, labels = month.abb) +
  scale_fill_manual(values = c("Winter" = "blue", "Spring" = "green", "Summer" = "yellow", "Fall" = "orange")) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 60, hjust = 1)
  )
```

@@

# Create the plot
ggplot(new_data, aes(x = month, y = predicted_outbreak, color = causative_agent_1)) +
  geom_rect(
    data = season_data, aes(xmin = start_month, xmax = end_month, ymin = ymin, ymax = ymax, fill = season),
    inherit.aes = FALSE, alpha = 0.2
  ) +
  geom_line() +
  geom_smooth(method = "loess", se = FALSE) +
  facet_wrap(~causative_agent_1, scales = "free_y") +
  labs(
    x = "Month",
    y = "Predicted Outbreak Count",
    fill = "Season",
    color = "Pathogen"
  ) +
  scale_x_continuous(breaks = 1:12, labels = month.abb) +
  scale_fill_manual(values = c("Winter" = "blue", "Spring" = "green", "Summer" = "yellow", "Fall" = "orange")) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 60, hjust = 1)
  )
```

@@

ggplot(combined_data, aes(x = month, y = count, color = source)) +
  geom_line(size = 1) +
  facet_wrap(~causative_agent_1, scales = "free_y") +
  labs(
    x = "Month",
    y = "Outbreak Count",
    color = "Source"
  ) +
  scale_x_continuous(breaks = 1:12, labels = month.abb) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 60, hjust = 1),
    legend.position = "bottom"
  )
```

@@

## More details about plot {#sec-plot-details}

The @fig-outbreakcount showcases the distribution of outbreak counts, revealing a heavily skewed pattern where most instances report lower outbreak counts. The majority of the data lies in the range of 0–2 outbreaks, indicating that many facilities or contexts experienced minimal disruptions. As the outbreak count increases, the frequency declines sharply, highlighting the rarity of high-count outbreaks. Notably, there are a few outlier cases, such as counts above 30 or even 50, which likely correspond to specific high-risk pathogens (e.g., COVID-19) or settings like long-term care homes. These extreme cases underline the need for targeted interventions in high-risk contexts.

In contrast, the @fig-outbreakcount aggregates all data without considering the influence of key predictors like pathogens or seasons. While it gives a general distribution of outbreak counts, it lacks the depth needed to examine contextual factors that affect outbreak patterns. This makes it less suitable for understanding the drivers of variation in outbreak frequencies across multiple factors. By visualizing outbreaks grouped by pathogen and season in @fig-count, we can identify trends or anomalies, such as seasonal peaks for specific pathogens, which are crucial for drawing actionable conclusions and targeting interventions.
```{r, fig.pos= 'H', fig.width=10, fig.height=6}
#| label: fig-outbreakcount
#| fig-cap: Distribution of outbreak Count
#| echo: false

ggplot(data, aes(x = outbreak_count)) +
  geom_bar(fill = "steelblue", color = "black", alpha = 0.8) +  # Add color and transparency
  labs(
    x = "Outbreak Count",
    y = "Frequency"
  ) +  # Add labels and title
  theme_minimal(base_size = 14) +  # Use a clean theme with larger text
  theme(
    plot.title = element_text(hjust = 0.5, face = "bold"),  # Center and bold the title
    axis.title.x = element_text(face = "bold"),  # Bold x-axis label
    axis.title.y = element_text(face = "bold")   # Bold y-axis label
  ) + 
  geom_text(stat = "count", aes(label = ..count..), vjust = -0.5, size = 4)

@@

par(mfrow = c(2, 3))
plot(poisson_season_time, which = 1:6)
```
